Avoiding tipping points in fisheries management through Gaussian Process Dynamic Programming

### 2014-10-28 manuscripts/manuscript.tex

We first describe the requirements of dynamic optimization for the
management of human intervention in natural resource systems. After that
we describe three parametric models for population dynamics and the GP
description of population dynamics.

@@

\[\mathbf{\mathrm{E}}_{X_t+1} \left[ V_{t+1}( X_{t+1}) | X_t, h_t \right] = \mathbf{\mathrm{E}}_{\theta|\mathrm{data}} \{ \mathbf{\mathrm{E}}_{f, Z | \theta, \mathrm{data}} \left[ V_{t+1}( Z f(X_t - h_t))  | x_t, h_t \right] \}\]

@@

@@

Then under the GP, given a vector of observed data $X_0$ the predicted
data $X_p$ obeys

\[f(X_p|X_o) \sim \mathcal{N}(E,C)\]
\[E = K(X_p, X_o) \left(K(X_o,X_o) - \sigma \mathbb{I} \right)  ^{-1} y\]
\[C = K(X_p, X_p) - K(X_p, X_o) K(X_o,X_o)^{-1} K(X_o, X_p)\]

@@

\[ K(x,y) = \exp\left(\frac{-(x-y)^2}{2 \ell^2} \right)\]

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_1-1.pdf}
\caption{Points show the training data of stock-size over time. Curves
show the posterior step-ahead predictions based on each of the estimated
models. Observe that all models are fitting the data reasonably well.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_2-1.pdf}
\caption{The inferred Gaussian process compared to the true process and
maximum-likelihood estimated process. We show the expected value for the
function $f$ under each model. Two standard deviations from the
estimated Gaussian process covariance with (light grey) and without
(darker grey) measurement error are also shown. The training data are
also shown as black points. The GP is conditioned on (0,0), shown as a
pseudo-data point.}
\end{figure}

We show the mean inferred state space dynamics of each model relative to
the true model used to generate the data in Figure 2, predicting the
relationship between observed stock size (x-axis) to the stock size
after recruitment the following year. Note that in contrast to the other
models shown, the expected Gaussian process corresponds to a
distribution of curves - as indicated by the gray band - which itself
has a mean shown in black. Parameter uncertainty (not shown) spreads out
the estimates further. The observed data from which each model is
estimated is also shown. The observations come from only a limited
region of state space corresponding to unharvested or weakly harvested
system. No observations occur at the theoretical optimum harvest rate or
near the tipping point. Figure S1 (see supplement) shows the performance
of the models outside the observed training data.

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_4-1.pdf}
\caption{The steady-state optimal policy (infinite boundary) calculated
under each model. Policies are shown in terms of target escapement,
$S_t$, as under models such as this a constant escapement policy is
expected to be optimal (Reed 1979).}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_5-1.pdf}
\caption{In the management context, GPDP outperforms approaches based on
parametric models. We show 100 replicate simulations of the stock
dynamics (eq 1) under the policies derived from each of the estimated
models, as well as the policy based on the exact underlying model.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_6-1.pdf}
\caption{Histograms of the realized net present value of the fishery
over a range of simulated data and resulting parameter estimates. For
each data set, the three models are estimated as described above. Values
plotted are the averages of a given policy over 100 replicate
simulations. Details and code provided in the supplement.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_7-1.pdf}
\caption{The effect of increasing noise or decreasing Allee threshold
levels on the net present value of the fishery when managed under the
GPDP, relative to managing under the true model (with known parameters).
Other than the focal parameter (stochasticity, Allee threshold), other
parameters are held fixed as above to illustrate this effect.}
\end{figure}

>>>>

We first describe the requirements of dynamic optimization for the
management of human intervention in natural resource systems. After that
we describe three parametric models for population dynamics and the
Gaussian Process (GP)\footnote{Note that we abbreviate Gaussian Process
  as GP, which refers only to the statistical model we use to
  approximate the population dynamics (as opposed to a familiar
  parametric model such as the Ricker equation), while we use the term
  GPDP, Gaussian Process Dynamic Programming, to refer to the \emph{use}
  of a GP as the underlying process model when solving a Dynamic
  Program, as we illustrate in this paper. Hence we will refer to the
  models as: GP, Ricker, Allen, etc, while we refer to the novel method
  we put forward here as GPDP.} description of population dynamics.

@@

\[\mathbf{\mathrm{E}}_{X_t+1} \left[ V_{t+1}( x_{t+1}) | X_t, h_t \right] = \mathbf{\mathrm{E}}_{\theta|\mathrm{data}} \{ \mathbf{\mathrm{E}}_{f, Z | \theta, \mathrm{data}} \left[ V_{t+1}( Z f(x_t - h_t))  | x_t, h_t \right] \}\]

@@

Throughout the presentation, we will refer to the use of the ``True''
model to denote solving for the optimal solution when the underlying
parameters \emph{are known without error}, and refer to the ``Allen''
model to denote the when these parameters have instead only been
estimated from the sample data.

@@

Then under the GP, given a vector of observed data $X_o$ the predicted
data $X_p$ obeys

\[f(X_p|X_o, X_{t+1}) \sim \mathcal{N}(E,C)\]
\[E = K(X_p, X_o) \left(K(X_o,X_o) - \sigma \mathbb{I} \right)^{-1} X_{t+1}\]
\[C = K(X_p, X_p) - K(X_p, X_o) \left(K(X_o,X_o) - \sigma \mathbb{I} \right)^{-1} K(X_o, X_p)\]

@@

\[ K(\mathbf{x},\mathbf{y}) = \exp\left(\frac{-|\mathbf{x}-\mathbf{y}|^2}{2 \ell^2} \right)\]

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_1-1.pdf}
\caption{Points show the training data of stock-size over time. Curves
show the posterior step-ahead predictions based on each of the estimated
models. Observe that all models are fitting the data reasonably well. In
the GP panel, black lines are means of the GP for random samples from
the posterior distributions, while grey bands around them reflect the
standard deviations of the GP distribution. (Recall that while given a
particular set of parameters, a parametric model gives just one curve,
the GP is always a distribution, whose mean we plot in black.)}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_2-1.pdf}
\caption{The inferred Gaussian process compared to the true process and
maximum-likelihood estimated process. We show the expected value for the
function $f$ under each model. Two standard deviations from the
estimated Gaussian process covariance with (light grey) and without
(darker grey) measurement error are also shown. The training data are
also shown as black points. The GP is conditioned on (0,0), shown as a
pseudo-data point, as explained in the text}
\end{figure}

We show the mean inferred state space dynamics of each model relative to
the true model used to generate the data in Figure 2, predicting the
relationship between observed stock size (x-axis) to the stock size
after recruitment the following year.\\In addition to the raw data,
Figure 2 illustrates that the GP is conditioned on going through the
point 0,0 without error. This conditioning on (0,0) is equivalent to
making the assumption that a species hitting 0 is extinct forever \&
without uncertainty, despite the lack of any data in the observed
sequence to justify this. The use of this assumption illustrates how the
GP approach can capture common-sense biology without having to assume
more explicit details about the dynamics at low population numbers that
have never been observed. Not that we need need not make this assumption
(perhaps there is immigration after all); and one could repeat the
entire analysis without this assumption.

Note that in contrast to the other models shown, the expected Gaussian
process corresponds to a distribution of curves - as indicated by the
gray band - which itself has a mean shown in black. Parameter
uncertainty (not shown) spreads out the estimates further. The observed
data from which each model is estimated is also shown. The observations
come from only a limited region of state space corresponding to
unharvested or weakly harvested system. No observations occur at the
theoretical optimum harvest rate or near the tipping point. Figure S1
(see supplement) shows the performance of the models outside the
observed training data.

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_3-1.pdf}
\caption{The steady-state optimal policy (infinite boundary) calculated
under each model. Policies are shown in terms of target escapement,
$S_t$, as under models such as this a constant escapement policy is
expected to be optimal (Reed 1979).}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_4-1.pdf}
\caption{In the management context, GPDP outperforms approaches based on
parametric models. We show 100 replicate simulations of the stock
dynamics (eq 1) under the policies derived from each of the estimated
models, as well as the policy based on the exact underlying model.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_5-1.pdf}
\caption{Histograms of the realized net present value of the fishery
over a range of simulated data and resulting parameter estimates. For
each data set, the three models are estimated as described above.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_6-1.pdf}
\caption{The effect of increasing noise or decreasing Allee threshold
levels on the net present value of the fishery when managed under the
GPDP, relative to managing under the true model (with known parameters).
Other than the focal parameter (stochasticity, Allee threshold), other
parameters are held fixed as above to illustrate this effect.}
\end{figure}
